Symbolic linearization of equations of motion of constrained multibody systems

### 2013-07-29 PetersonGedeHubbard2012.tex
Work on example.

With these definitions established, Kane's method can be applied.
The output of Kane's method is the equation
\begin{align}
\label{eq:kanes_eq}
\mathbf{F} + \mathbf{F}^* = \mathbf{0}
\end{align}
or, the sum of the generalized active forces and generalized inertial forces is
0.
If Newton's second law were written as $f - ma = 0$, the $\mathbf{F}$ is the
multibody generalization of $f$ and $\mathbf{F}^*$ is the multibody
generalization of $-ma$.
These terms are created by taking the dot product of the partial
velocities/angular-velocities, as defined in (\ref{eq:particle_translational})
- (\ref{eq:rb_rotational_frames}), with the resultant forces/torques ($\bar{R}$/
$\bar{T}$) and the inertial forces/torques ($\bar{R}^*$/$\bar{T}^*$).

@@

\section{Dynamics Example}
To illustrate Kane's method in the presence of configuration and velocity
constraints, we consider a thin disc $C$ rolling without slip on a horizontal
plane. The dextral unit vectors $\bm{c}_x, \bm{c}_y, \bm{c}_z$ are fixed to $C$
with $\bm{c}_y$ perpendicular to the disc plane. We take $r$ and $m$ to be the
disc radius and mass, respectively, and assume the mass distribution of a thin
disc so that $I_{xx}=I_{zz}=mr^2/4$ and $I_{yy} = mr^2$. We take the inertial
(Newtonian) frame $N$ to be similarly equipped with dextral unit vectors
$\bm{n}_x, \bm{n}_y, \bm{n}_z$, with $\bm{n}_z$ perpendicular to the ground
plane and aligned with local gravitational field ($\bm{n}_z$ points downwards).
To orient the disc, we first align the $C$ with $N$, then apply successive body
fixed ZYX rotations: yaw $q_1$, roll $q_2$, and spin $q_3$.  We denote by $A$
and $B$ the two frames associated with the first two intermediate rotations in
the sequence; we refer to these as the instantaneous yaw and lean frames,
respectively. The unit vector $\bm{b}_z$ fixed in the $B$ frame is parallel to
the disc plane and aligned with the position vector from the disc center $C^*$
to the lowest point of the disc $P^*$. Traditionally, to locate the center of
the disc $C^*$ relative to the inertial origin $N^*$, two coordinates are used
to locate the contact point in the ground plane; those two coordinates plus the
lean angle $q_2$ determine the location of the center of the disc. The
advantage of this approach is that the choice of coordinates guarantees the
disc remain in contact with the ground plane. To illustrate how dependent
coordinates must be accounted for when linearizing the equations of motion, we
purposefully locate the center of the disc $C^*$ relative to the inertial
origin $N^*$ with a non-minimal choice of coordinates
\begin{equation*}
  \bm{r}^{C^*} = q_4 \bm{n}_x + q_5 \bm{n}_y + q_6 \bm{n}_z
\end{equation*}
which results in the configuration constraint
\begin{equation}
 r\cos{q_2} + q_6 = 0
\end{equation}
which must be satisfied for the disc to remain in contact with the ground.
While this constraint is easily avoided by appropriate choice of coordinates,
it serves to illustrate considerations which must be made when analyzing
systems where there is no obvious way to eliminate configuration constraints.

Again for the purpose of illustrating how velocity constraints must be
accounted for when obtaining linearized equations of motion, we define
the angular velocity of $C$ and the velocity of $C^*$ relative to $N$ using a
non-minimal choice of generalized speeds
\begin{align}
  \label{w_u}
  \bm{\omega}^C &= u_1 \bm{c}_x + u_2 \bm{c}_y + u_3 \bm{c}_z \\
  \label{v_u}
  \bm{v}^{C^*} &= u_4 \bm{c}_x + u_5 \bm{c}_y + u_6 \bm{c}_z
\end{align}
Given these definitions, the velocity of the lowest point of the disc $P^*$ is
\begin{align*}
  \bm{v}^{P^*} &= \bm{v}^{C^*} + \bm{\omega}^{C} \times r^{P^*/C^*} \\
  &= (r u_2 \cos q_3  + u_4) \bm{c}_x
   + (-r (u_1 \cos q_3  + u_3 \sin q_3) + u_5) \bm{c}_y \\
   &+ (r u_2 \sin q_3  + u_6) \bm{c}_z
\end{align*}
where we have made use of the fact that $\bm{r}^{P^*/C^*} = r\bm{b}_z$.
Under the assumption of pure rolling this immediately yields three velocity
constraints
\begin{align}
  r u_2 \cos q_3  + u_4 &= 0\\
  -r (u_1 \cos q_3  + u_3 \sin q_3) + u_5 &= 0\\
  r u_2 \sin q_3  + u_6 &= 0
\end{align}
These equations are linear in the $u_i$'s, nonlinear in the $q_i$'s, involve
geometric system parameters (but not mass or inertial parameters), and do not
explicitly involve $\dot{q}_i$'s. This structure will be taken advantage of
subsequently.

@@

The kinematic differential equations relate the time derivatives of the
coordinates and the generalized speeds. From the addition theorem for angular
velocity, we may write the angular velocity of $C$ as
\begin{equation}
  \label{w_qdot}
  \bm{\omega}^C = \dot{q}_1 \bm{n}_z + \dot{q}_2 \bm{a}_x + \dot{q}_3 \bm{b}_y
\end{equation}
while time differentiating $r^{C^*}$ in $N$ yields
\begin{equation}
  \label{v_qdot}
  \bm{v}^{C^*} = \dot{q}_4 \bm{n}_x + \dot{q}_5 \bm{n}_y + \dot{q}_6 \bm{n}_z
\end{equation}
Equating (\ref{w_u}) with (\ref{w_qdot}) and (\ref{v_u}) with (\ref{v_qdot}),
and resolving these vector equations into the $B$ frame, and rearranging so
that all terms appear to the left of the equality sign, we obtain the kinematic
differential equations
\begin{align}
  \dot{q}_2 - u_1 \cos q_3  - u_3 \sin q_3 &= 0 \\
  \dot{q}_1 \sin q_2 + \dot{q}_3 - u_2 &= 0 \\
  \dot{q}_1 \cos q_2 + u_1 \sin q_3  - u_3 \cos q_3  &= 0 \\
  \dot{q}_4 \cos q_1 + \dot{q}_5 \sin q_1 -u_4 \cos q_3 - u_6 \sin q_3 &= 0 \\
 -\dot{q}_4 \sin q_1 \cos q_2 + \dot{q}_5 \cos q_1 \cos q_2 + \dot{q}_6 \sin q_2 - u_5 &= 0 \\
  \dot{q}_4 \sin q_1 \sin q_2 - \dot{q}_5 \sin q_2 \cos q_1 + \dot{q}_6 \cos q_2 + u_4 \sin q_3 - u_6 \cos q_3 &= 0
\end{align}
Before formulating the dynamic equations, we

>>>>

With these definitions established, Kane's method can be applied.
The output of Kane's method, in vector form, is the equation
\begin{align}
\label{eq:kanes_eq}
\mathbf{F} + \mathbf{F}^* = \mathbf{0}
\end{align}
or, the sum of the generalized active forces and generalized inertial forces is
0.
If Newton's second law were written as $f - ma = 0$, the $\mathbf{F}$ is the
multibody generalization of $f$ and $\mathbf{F}^*$ is the multibody
generalization of $-ma$.
These terms are created by taking the dot product of the partial
velocities/angular-velocities, as defined in (\ref{eq:particle_translational})
- (\ref{eq:rb_rotational_frames}), with the resultant forces/torques ($\bar{R}$/
$\bar{T}$) and the inertial forces/torques ($\bar{R}^*$/$\bar{T}^*$).

@@

\section{Dynamics Example}
To illustrate Kane's method in the presence of configuration and velocity
constraints, we consider a thin disc $C$ rolling without slip on a horizontal
plane. The dextral unit vectors $\hat{c}_x, \hat{c}_y, \hat{c}_z$ are fixed to $C$
with $\hat{c}_y$ perpendicular to the disc plane. We take $r$ and $m$ to be the
disc radius and mass, respectively, and assume the mass distribution of a thin
disc so that $I_{xx}=I_{zz}=mr^2/4$ and $I_{yy} = mr^2/2$. We take the inertial
(Newtonian) frame $N$ to be similarly equipped with dextral unit vectors
$\hat{n}_x, \hat{n}_y, \hat{n}_z$, with $\hat{n}_z$ perpendicular to the ground
plane and aligned with local gravitational field ($\hat{n}_z$ points downwards).
To orient the disc, we first align the $C$ with $N$, then apply successive body
fixed ZYX rotations: yaw $q_1$, roll $q_2$, and spin $q_3$.  We denote by $A$
and $B$ the two frames associated with the first two intermediate rotations in
the sequence; we refer to these as the instantaneous yaw and lean frames,
respectively. The unit vector $\hat{b}_z$ fixed in the $B$ frame is parallel to
the disc plane and aligned with the position vector from the disc center $C^*$
to the lowest point of the disc $P^*$. Traditionally, to locate the center of
the disc $C^*$ relative to the inertial origin $N^*$, two coordinates are used
to locate the contact point in the ground plane; those two coordinates plus the
lean angle $q_2$ determine the location of the center of the disc. The
advantage of this approach is that the choice of coordinates guarantees the
disc remain in contact with the ground plane. To illustrate how dependent
coordinates must be accounted for when linearizing the equations of motion, we
purposefully locate the center of the disc $C^*$ relative to the inertial
origin $N^*$ with a non-minimal choice of coordinates
\begin{equation*}
  \bar{r}^{C^*} = q_4 \hat{n}_x + q_5 \hat{n}_y + q_6 \hat{n}_z
\end{equation*}
which results in the configuration constraint
\begin{equation}
  r\cos{q_2} + q_6 = 0
\end{equation}
which must be satisfied for the disc to remain in contact with the ground.
While this constraint is easily avoided by appropriate choice of coordinates,
it serves to illustrate considerations which must be made when analyzing
systems where there is no obvious way to eliminate configuration constraints.

Again for the purpose of illustrating how velocity constraints must be
accounted for when obtaining linearized equations of motion, we define
the angular velocity of $C$ and the velocity of $C^*$ relative to $N$ using a
non-minimal choice of generalized speeds
\begin{align}
  \label{w_u}
  {^N}\bar{\omega}^C &= u_1 \hat{c}_x + u_2 \hat{c}_y + u_3 \hat{c}_z \\
  \label{v_u}
  {^N}\bar{v}^{C^*} &= u_4 \hat{c}_x + u_5 \hat{c}_y + u_6 \hat{c}_z
\end{align}
It should be immediately obvious from these defintions that the partial angular
velocity and partial velocity of $C^*$, respectively, are
\begin{align}
  {^N}\bar{\omega}^C_u &= [\hat{c}_x, \hat{c}_y, \hat{c}_z, 0, 0, 0] \\
  {^N}\bar{v}^{C^*}_u &= [0, 0, 0, \hat{c}_x, \hat{c}_y, \hat{c}_z]
\end{align}
where we present the partial velocities as a $1\times6$ matrix whose entries
are populated with standard 3-vectors. Given these definitions, the velocity of
the lowest point of the disc $P^*$ is
\begin{align*}
  \bm{v}^{P^*} &= \bm{v}^{C^*} + \bm{\omega}^{C} \times r^{P^*/C^*} \\
  &= (r u_2 \cos q_3  + u_4) \bm{c}_x
   + (-r (u_1 \cos q_3  + u_3 \sin q_3) + u_5) \bm{c}_y \\
   &+ (r u_2 \sin q_3  + u_6) \bm{c}_z
\end{align*}
where we have made use of the fact that $\bm{r}^{P^*/C^*} = r\bm{b}_z$.
Under the assumption of pure rolling this immediately yields three velocity
constraints
\begin{align}
  r u_2 \cos q_3  + u_4 &= 0\\
  -r (u_1 \cos q_3  + u_3 \sin q_3) + u_5 &= 0\\
  r u_2 \sin q_3  + u_6 &= 0
\end{align}
These equations are linear in the $u_i$'s, nonlinear in the $q_i$'s, involve
geometric system parameters (but not mass or inertial parameters), and do not
explicitly involve $\dot{q}_i$'s. This structure will be taken advantage of
subsequently.

@@

The kinematic differential equations relate the time derivatives of the
coordinates and the generalized speeds. From the addition theorem for angular
velocity, we may write the angular velocity of $C$ as
\begin{equation}
  \label{w_qdot}
  {^N}\bar{\omega}^C = \dot{q}_1 \hat{n}_z + \dot{q}_2 \hat{a}_x + \dot{q}_3 \hat{b}_y
\end{equation}
while time differentiating $r^{C^*}$ in $N$ yields
\begin{equation}
  \label{v_qdot}
  \bar{v}^{C^*} = \dot{q}_4 \hat{n}_x + \dot{q}_5 \hat{n}_y + \dot{q}_6 \hat{n}_z
\end{equation}
Equating (\ref{w_u}) with (\ref{w_qdot}) and (\ref{v_u}) with (\ref{v_qdot}),
and resolving these vector equations into the $B$ frame, and rearranging so
that all terms appear to the left of the equality sign, we obtain
\begin{align}
  \dot{q}_2 - u_1 \cos q_3  - u_3 \sin q_3 &= 0 \\
  \dot{q}_1 \sin q_2 + \dot{q}_3 - u_2 &= 0 \\
  \dot{q}_1 \cos q_2 + u_1 \sin q_3  - u_3 \cos q_3  &= 0 \\
  \dot{q}_4 \cos q_1 + \dot{q}_5 \sin q_1 -u_4 \cos q_3 - u_6 \sin q_3 &= 0 \\
 -\dot{q}_4 \sin q_1 \cos q_2 + \dot{q}_5 \cos q_1 \cos q_2 + \dot{q}_6 \sin q_2 - u_5 &= 0 \\
  \dot{q}_4 \sin q_1 \sin q_2 - \dot{q}_5 \sin q_2 \cos q_1 + \dot{q}_6 \cos q_2 + u_4 \sin q_3 - u_6 \cos q_3 &= 0
\end{align}
The angular acceleration of $C$ and the translation acceleration of $C^*$,
relative to $N$, is
\begin{align}
  {^N}\bar{\alpha}^{C} &= \dot{u}_1 \hat{c}_x + \dot{u}_2 \hat{c}_y + \dot{u}_3 \hat{c}_z  \\
  {^N}\bar{a}^{C^*} &= (u_2 u_6 - u_3 u_5 + \dot{u}_4) \hat{c}_x + (u_3 u_4 - u_1 u_6 +
+ \dot{u}_5) \hat{c}_y \notag\\&+ (u_1 u_5 - u_2 u_4 + \dot{u}_6) \hat{c}_z
\end{align}
